---
title: "Neural network visualisation"
output: html_document
---


The goal of this assignement is the visualisation of neural networks. This plotting function is based on neural networks with the nnet or neuralnet functions, it also works with numeric weight functions.

#Primary functions

##tests functions
This function tests the input parameters 
```{r}
tests<-function(model,struct){
  ###
  #tests the input parameters 
  #
  #PARAMETERS
  #----------
  #model: neural network object or numeric vector of weights
  #struct:numeric value of length three 
  #indicating network architecture(no. nodes for input,  hidden,output)
  ###
  
  if('numeric' %in% class(model)){
    if(is.null(struct)) stop('Three-element vector required for struct')
    if(length(model) != ((struct[1]*struct[2]+struct[2]*struct[3])+(struct[3]+struct[2])))
      stop('Incorrect length of weight matrix for given network structure')
  }
  if('train' %in% class(model)){
    if('nnet' %in% class(model$finalModel)){
      model<-model$finalModel
      warning('Using best nnet model from train output')
    }
    else stop('Only nnet method can be used with train object')
  }
}
```


##Weight extraction
The following function extracts the weights from the models 

```{r}
weight_extraction<-function(model,nid,width,struct.out=struct){
  ###
  #Returns the weight matrix from the neural network model,output is list
  #
  #PARAMETERS
  #----------
  #model: neural network object or numeric vector of weights
  #Width: The maximum width of connection lines 
  #struct.out:numeric value of length three indicating network architecture(no. nodes for input,  hidden,output), required only if mod.in is a numeric vector 
  ###
  require(reshape)
  require(scales)
  
  #first case: model is a weight matrix  
  if('numeric' %in% class(model)){
      wts<-model
  }
  #Second case: neuralnet function
  if('nn' %in% class(model)){
    struct.out<-unlist(lapply(model$weights[[1]],ncol))
    struct.out<-struct.out[-length(struct.out)]
    struct.out<-c(
    	length(model$model.list$variables),
    	struct.out,
    	length(model$model.list$response)
    	)    		
    wts<-unlist(model$weights[[1]])   
    }
  #Third case: nnet package
  if('nnet' %in% class(model)){
    struct.out<-model$n
    wts<-model$wts
  }
  
  #rescale
  if(nid) wts<-rescale(abs(wts),c(1,width))
  
  #convert wts to list with appropriate names 
  hid.struct<-struct.out[-c(length(struct.out))]
  row.nms<-NULL
  for(i in 1:length(hid.struct)){
    if(is.na(hid.struct[i+1])) break
    row.nms<-c(row.nms,rep(paste('hidden',i,seq(1:hid.struct[i+1])),each=1+hid.struct[i]))
  }
  row.nms<-c(
    row.nms,
    rep(paste('out',seq(1:struct.out[length(struct.out)])),each=1+struct.out[length(struct.out)-1])
  )
  out.ls<-data.frame(wts,row.nms)
  out.ls$row.nms<-factor(row.nms,levels=unique(row.nms),labels=unique(row.nms))
  out.ls<-split(out.ls$wts,f=out.ls$row.nms)
    
  assign('struct',struct.out,envir=.GlobalEnv)

  out.ls 
    
  }
```



##Variables names
```{r}
Variable_names<-function(model,struct,x.lab,y.lab){
  #get variable names from model object
  #change to user input if supplied
  #
  #PARAMETERS 
  #----------
  #model: neural network object or numeric vector of weights
  #struct: the architecture of the neural network 
  #x.lab/y.lab: user input for variable names 
  
  if('numeric' %in% class(model)){
    x.names<<-paste0(rep('X',struct[1]),seq(1:struct[1]))
    y.names<<-paste0(rep('Y',struct[3]),seq(1:struct[3]))
  }
  if('nn' %in% class(model)){
    x.names<<-model$model.list$variables
    y.names<<-model$model.list$respons
  }
  if('xNames' %in% names(model)){
    x.names<<-model$xNames
    y.names<<-attr(terms(model),'factor')
    y.names<<-row.names(y.names)[!row.names(y.names) %in% x.names]
  }
  if(!'xNames' %in% names(model) & 'nnet' %in% class(model)){
    if(is.null(model$call$formula)){
      x.names<<-colnames(eval(model$call$x))
      y.names<<-colnames(eval(model$call$y))
    }
    else{
      forms<<-eval(model$call$formula)
      x.names<<-model$coefnames
      facts<<-attr(terms(model),'factors')
      y.check<<-model$fitted
      if(ncol(y.check)>1) y.names<<-colnames(y.check)
      else y.names<<-as.character(forms)[2]
    } 
  }
  #change variables names to user sub 
  if(!is.null(x.lab)){
    if(length(x.names) != length(x.lab)) 
      stop('x.lab length not equal to number of input variables')
    else x.names<<-x.lab
  }
  if(!is.null(y.lab)){
    if(length(y.names) != length(y.lab)) 
      stop('y.lab length not equal to number of output variables')
    else y.names<<-y.lab
  }
}
```

## Plotting functions
```{r}
get.ys<-function(lyr){
  #function for getting y locations for input, hidden, output layers
  #input is integer value from 'struct'
  spacing<-diff(c(0*diff(y.range),0.9*diff(y.range)))/max(struct)
  
  		seq(0.5*(diff(y.range)+spacing*(lyr-1)),0.5*(diff(y.range)-spacing*(lyr-1)),
        length=lyr)
  }
```

```{r}
layer.points<-function(layer,x.loc,layer.name,cex){
  #function for plotting nodes
  #
  #PARAMETERS
  #---------
  #cex: numeric value indicating size of text labels
  #layer: specifies which layer, integer from 'struct'
  #x.loc:indicates x location for layer, integer from 'layer.x'
  #layer.name: is string indicating text to put in node
  x<-rep(x.loc*diff(x.range),layer)
  y<-get.ys(layer)
  points(x,y,pch=21,cex=circle.cex,col=in.col,bg=bord.col)
  text(x,y,paste(layer.name,1:layer,sep=''),cex=cex.val)
  if(layer.name=='I' & var.labs) text(x-line.stag*diff(x.range),y,x.names,pos=2,cex=cex.val)      
  if(layer.name=='O' & var.labs) text(x+line.stag*diff(x.range),y,y.names,pos=4,cex=cex.val)
  }

```

```{r}
bias.points<-function(bias.x,bias.y,layer.name,cex,...){
  #function for plotting bias points
  #'bias.x' is vector of values for x locations
  #'bias.y' is vector for y location
  #'layer.name' is  string indicating text to put in node
  for(val in 1:length(bias.x)){
    points(
      diff(x.range)*bias.x[val],
      bias.y*diff(y.range),
      pch=21,col=in.col,bg=bord.col,cex=circle.cex
    )
     text(
        diff(x.range)*bias.x[val],
        bias.y*diff(y.range),
        paste(layer.name,val,sep=''),
        cex=cex.val
      )
  }
}
```

```{r}
layer.lines<-function(model,h.layer,layer1=1,layer2=2,out.layer=F,width,nid,pos.col,
                        neg.col,...){
  #function creates lines colored by direction and width as proportion of magnitude
    
    x0<-rep(layer.x[layer1]*diff(x.range)+line.stag*diff(x.range),struct[layer1])
    x1<-rep(layer.x[layer2]*diff(x.range)-line.stag*diff(x.range),struct[layer1])
    
    if(out.layer==T){
      
      y0<-get.ys(struct[layer1])
      y1<-rep(get.ys(struct[layer2])[h.layer],struct[layer1])
      src.str<-paste('out',h.layer)
      
      wts<-weight_extraction(model,nid=F,width)
      wts<-wts[grep(src.str,names(wts))][[1]][-1]
      wts.rs<-weight_extraction(model,nid=T,width)
      wts.rs<-wts.rs[grep(src.str,names(wts.rs))][[1]][-1]
      
      cols<-rep(pos.col,struct[layer1])
      cols[wts<0]<-neg.col
      
      if(nid) segments(x0,y0,x1,y1,col=cols,lwd=wts.rs)
      else segments(x0,y0,x1,y1)
      
    }
    
    else{
      
      y0<-rep(get.ys(struct[layer1])[h.layer],struct[2])
      y1<-get.ys(struct[layer2])
      src.str<-paste('hidden',layer1)
      
      wts<-weight_extraction(model,nid=F,width)
      wts<-unlist(lapply(wts[grep(src.str,names(wts))],function(x) x[h.layer+1]))
      wts.rs<-weight_extraction(model,nid=T,width)
      wts.rs<-unlist(lapply(wts.rs[grep(src.str,names(wts.rs))],function(x) x[h.layer+1]))
      
      cols<-rep(pos.col,struct[layer2])
      cols[wts<0]<-neg.col
      
      segments(x0,y0,x1,y1,col=cols,lwd=wts.rs)
    }
    
  }

```

```{r}
bias.lines<-function(bias.x,model,nid,width,pos.col,neg.col,circle.cex=5,...){
  x.range<-c(0,100)
  y.range<-c(0,100)
  #these are all proportions from 0-1
  line.stag<-0.011*circle.cex/2
  layer.x<-seq(0.17,0.9,length=length(struct))
  bias.x<-layer.x[-length(layer.x)]+diff(layer.x)/2
  bias.y<-0.95
  circle.cex<-circle.cex
  all.out<-1:struct[length(struct)]
    
  for(val in 1:length(bias.x)){
    wts<-weight_extraction(model,nid=F,width)
    wts.rs<-weight_extraction(model,nid=T,width)
    if(val != length(bias.x)){
      wts<-wts[grep('out',names(wts),invert=T)]
      wts.rs<-wts.rs[grep('out',names(wts.rs),invert=T)]
    	sel.val<-grep(val,substr(names(wts.rs),8,8))
    	wts<-wts[sel.val]
    	wts.rs<-wts.rs[sel.val]
    }
    
    else{
      wts<-wts[grep('out',names(wts))]
      wts.rs<-wts.rs[grep('out',names(wts.rs))]
    }
      
    cols<-rep(pos.col,length(wts))
    cols[unlist(lapply(wts,function(x) x[1]))<0]<-neg.col
    wts.rs<-unlist(lapply(wts.rs,function(x) x[1]))
      
    if(nid==F){
      wts.rs<-rep(1,struct[val+1])
      cols<-rep('black',struct[val+1])
    }
  
    if(val != length(bias.x)){
      segments(
        rep(diff(x.range)*bias.x[val]+diff(x.range)*line.stag,struct[val+1]),
        rep(bias.y*diff(y.range),struct[val+1]),
        rep(diff(x.range)*layer.x[val+1]-diff(x.range)*line.stag,struct[val+1]),
        get.ys(struct[val+1]),
        lwd=wts.rs,
        col=cols
      )
    }
      
    else{
      segments(
        rep(diff(x.range)*bias.x[val]+diff(x.range)*line.stag,struct[val+1]),
        rep(bias.y*diff(y.range),struct[val+1]),
        rep(diff(x.range)*layer.x[val+1]-diff(x.range)*line.stag,struct[val+1]),
        get.ys(struct[val+1])[all.out],
        lwd=wts.rs[all.out],
        col=cols[all.out]
      )
    }
      
  }
}
```




#Main program 
```{r}
neural_network_plot<- function(model,struct1=NULL,nid=T,bias=T,width=5,circle.cex=5,
                     var.labs=T,x.lab=NULL,y.lab=NULL,cex.val=1, wts.only=F,alpha.val=1,
                     circle.col='lightblue',pos.col='blue',neg.col='grey', ...){
  require(scales)
  cex.val<<-cex.val
  var.labs<<-var.labs
  #1) Tests:
  tests(model,struct=struct1)
  
  #2) Weight extraction:
  wts<-weight_extraction(model,nid,width,struct.out=struct1)
  if(wts.only) return(wts)
  #3)Colors:
  ##circle colors for input, if desired, must be two-vector list, first vector is for input layer
  if(is.list(circle.col)){
                    circle.col.inp<-circle.col[[1]]
                    circle.col<-circle.col[[2]]
                    }
  else circle.col.inp<-circle.col
  
  #4) initiate plotting
  x.range<<-c(0,100)
  y.range<<-c(0,100)
  #these are all proportions from 0-1
  line.stag<<-0.011*circle.cex/2
  layer.x<<-seq(0.17,0.9,length=length(struct))
  bias.x<<-layer.x[-length(layer.x)]+diff(layer.x)/2
  bias.y<<-0.95
  circle.cex<<-circle.cex
  
  #5) Variable names 
  Variable_names(model,struct,x.lab,y.lab)
  
  #6) initiate plot 
  plot(x.range,y.range,type='n',axes=F,ylab='',xlab='',...)
  
  #7) biais lines 
  if(bias) bias.lines(bias.x,model,nid=nid,width=width,pos.col=alpha(pos.col,alpha.val),
                      neg.col=alpha(neg.col,alpha.val))
  
  
  #8)Layer lines
  mapply(
      function(x) layer.lines(model,x,layer1=1,layer2=2,nid=nid,width=width,
                              pos.col=alpha(pos.col,alpha.val),neg.col=alpha(neg.col,alpha.val)),
      1:struct[1]
  )
  
  #9): connections between hidden layers
  lays<-split(c(1,rep(2:(length(struct)-1),each=2),length(struct)),
              f=rep(1:(length(struct)-1),each=2))
  lays<-lays[-c(1,(length(struct)-1))]
  for(lay in lays){
    for(node in 1:struct[lay[1]]){
      layer.lines(model,node,layer1=lay[1],layer2=lay[2],nid=nid,width=width,
                  pos.col=alpha(pos.col,alpha.val),neg.col=alpha(neg.col,alpha.val))
    }
  }
  
  #10) lines for hidden-output
  mapply(
    function(x) layer.lines(model,x,layer1=length(struct)-1,layer2=length(struct),out.layer=T,nid=nid,
                            width=width,
                            pos.col=alpha(pos.col,alpha.val),neg.col=alpha(neg.col,alpha.val)),
    1:struct[length(struct)]
  )
  
  #use functions to plot nodes
  for(i in 1:length(struct)){
    bord.col<<-circle.col
    in.col<<-bord.col
    layer.name<-'H'
    if(i==1) { layer.name<-'I'; in.col<<-bord.col<<-circle.col.inp}
    if(i==length(struct)) layer.name<-'O'
    layer.points(struct[i],layer.x[i],layer.name)
    }

  if(bias) bias.points(bias.x,bias.y,'B')
}
```





#TEST:

```{r}
library(clusterGeneration)
 
seed.val<-2
set.seed(seed.val)
 
num.vars<-8
num.obs<-1000
 
#input variables
cov.mat<-genPositiveDefMat(num.vars,covMethod=c("unifcorrmat"))$Sigma
rand.vars<-mvrnorm(num.obs,rep(0,num.vars),Sigma=cov.mat)
 
#output variables
parms<-runif(num.vars,-10,10)
y1<-rand.vars %*% matrix(parms) + rnorm(num.obs,sd=20)
parms2<-runif(num.vars,-10,10)
y2<-rand.vars %*% matrix(parms2) + rnorm(num.obs,sd=20)
 
#final datasets
rand.vars<-data.frame(rand.vars)
resp<-data.frame(y1,y2)
names(resp)<-c('Y1','Y2')
dat.in<-data.frame(resp,rand.vars)
```

```{r}
#nnet function from nnet package
library(nnet)
set.seed(seed.val)
mod1<-nnet(rand.vars,resp,data=dat.in,size=6,linout=T)
```
```{r}
neural_network_plot(mod1)
```



